---
title: "Practicum Overview"
author: "Senetcky, Alexander; Archambault, Gary"
date: '2022-09-27'
output:
   html_document:
    toc: true # table of content true
    toc_float: true
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objectives  

Over the course of this practicum it is our hopes that our student will be able to
demonstrate proficiency or an appropriate level of understanding for the following:  

## Knowledge based skills

* A working knowledge of the statistical programming language, R, and its
family of packages known as the 
[tidyverse](https://rviews.rstudio.com/2017/06/08/what-is-the-tidyverse/)

* A working knowledge of the 
[R Studio Integrated Development Environment (IDE)](https://www.rstudio.com/products/rstudio/)

* A working knowledge of using R to query Star-model based data housed within
SQL databases 

* A working knowledge of Connecticut's 
[Open Data Portal](https://data.ct.gov/)(ODP)

* A basic understanding of Devops and AGILE workflows within the Azure DevOps tool

* A basic understanding of Git, a distributed source control

## Project based Outcomes 

* An ODP data product that will more easily serve the public a tabular version 
of the data found on DPH's [Infectious Diseases Statistics page](https://portal.ct.gov/DPH/Epidemiology-and-Emerging-Infections/Infectious-Diseases-Statistics)



# Timeline

## Getting set up
1. Getting Access to our developer environment and star model
1. R + R Studio Bootcamp
1. Git crashcourse
1. R usescases at the DPH

## Project Development
1. Primer on working in Azure Devops - primarily git focused
1. Overview of the data on [Infectious Diseases Statistics page](https://portal.ct.gov/DPH/Epidemiology-and-Emerging-Infections/Infectious-Diseases-Statistics)
    1. Pros and Cons - how do we think we can make it better
1. Pulling the data from the star model
1. Comparing the data from the star model to previous public releases
1. Creating a script that can be the basis of our automation that will move data
from the star model to the ODP for public consumption


# Suggested Reading 1

## R Introduction
[R for Data Science](https://r4ds.had.co.nz/)
Chapters/ sections, 1, 2, 3.1-3.1.1


# R and the ODP

## Getting Data from the ODP
I find the best way to learn R, or any language is to start with a problem, break it 
in to pieces and then start tackling from there.  One of the earliest hurdles
an epidemiologist might have to overcome involves reading data into the environment.
Usually there a few options and ways to go about this per a language. SAS and R Studio
both have a point and click gui to read-in data, and they both have options to 
ingest data programmatically as well. R is generally easier to do this because it is 
less strict on data type and length, but a good programmer will generally enforce 
these things anyway.  Today we'll be doing something a little different, since one
of the goals is to learn the Open Data Portal, we are going to play with data that
is stored on the open web and pull it down using an Application Programming Interface
or API.
  
  
## Installing RSocrata Package
Below is a code chunk showing the installation of the RSocrata Package. For your first
time installing this, you will want to erase the # and run that line to install the
package.  
  
Below that is the library statement that invokes the use of that package's
functions for your workflow. We Will make heavy use of the function, _read.socrata()_.
You can see it in action below, where I pull in publicly released CT Monkey Pox
Age group data. The ODP let's users do an API callout to pull data into memory.  The 
URL listed there is the unique API webaddress for that 
[dataset](https://data.ct.gov/Health-and-Human-Services/CT_CaseReportableDisease_AgeGroup/bg8t-bfyp).

```{r installing_RSocrata, }
#install.packages("RSocrata")
library(RSocrata)

#install.packages("tidyverse")
library(tidyverse)


monkey_pox_agegroups <- 
  RSocrata::read.socrata(
    url = "https://data.ct.gov/resource/bg8t-bfyp.json"
  ) %>% 
  as_tibble()


print(monkey_pox_agegroups)

```


## Suggested Reading 2
### R Visualizations
[R for Data Science](https://r4ds.had.co.nz/)
Chapters/ sections 3.2-3.4, skim 3.5, 3.6, skim 3.7-3.10

## Plotting ODP Monkey Pox Data

Armed with the reading above, and the _read.socrata()_ example, please attempt to do
the following:

* Create  a new .R script
* Download the MPOX state 
[data](https://data.ct.gov/Health-and-Human-Services/CT_CaseReportableDisease_State/qhtt-czu2)
using the URL "https://data.ct.gov/resource/qhtt-czu2.json" and assign it to a tibble
* Using the ReportPeriodStart variable for the x-axis and a count on the y axis, 
please create a bar chart using ggplot2 package 
(which is part of the tidyverse package) and _geom_bar()_

```{r eval=FALSE, include=FALSE}
monkey_pox_state_data <- 
  RSocrata::read.socrata(
    url = "https://data.ct.gov/resource/qhtt-czu2.json"
  ) %>% 
  as_tibble()


monkey_pox_state_data %>% 
  ggplot() +
  geom_bar(aes(reportperiodstart))

```


# DPLYR (The bread and butter of the tidyverse)

## Suggested Reading

* [Quick Overview](https://dplyr.tidyverse.org/) Just read the overview and check out
the Cheat Sheet

* [Introduction to dplyr](https://dplyr.tidyverse.org/articles/dplyr.html) This whole
thing is a good overview. Focusing on _filter_, _select_, _rename_, _mutate_ and 
_summarise_ is a good use of time though.


## Approaching a table with SAS and dplyr 

SAS and R approach similar issues from different points of view. I think both are 
important to know. Languages come and go, and tools can be swapped for others, but
keeping in mind how one language might tackle a problem and then being able to apply
it to novel roadblocks or new languages is always helpful. Long story short, in 
public health you will always strapped for time and/or resources, thinking critically
and being able to break a problem down into miniature steps like you would in any
language is possibly the most important lesson of all.  
  
In general a SAS workflow might center around a few workhouse functions with many 
different arguments, whereas R in the tidyverse, will focus around smaller functions
that has less options, but by doing one thing, you can chain them together in any 
order.  

SAS PROC FREQ is a good example, out of the box it is a handy little function for 
looking at counts across groups in data.  Using our monkey_pox_agegroups table
as an example you might type something like the following in SAS:

(bear with me, my SAS is geting rusty :D)


PROC FREQ data = monkey_pox_agegroups;  
Tables agegroup;  
Run;  

This would give you a count, a groupwise percent of the whole, a cumulative count, 
and a cumulative percent.  Not bad for a few lines. If we try to recreate that in R
using the basic dplyr verbs, it might be a bit more verbose.  However, you might not
need all the extras, or as you become more advanced and you settle into a more 
standard approach in your day to day, you might try your hand at creating 
your own functions and/or starting piping functions into other functions using some 
other packages.  It's in that sweet spot where you start getting a return on your 
investment.  But for now, let's use recreating proc freq as a showcase for 
showing off dplyr verbs.


```{r, dplyr showcase}
#let's take a look at the table structure again

glimpse(monkey_pox_agegroups)


#base R might use the str() function instead
str(monkey_pox_agegroups) #think structure
 

#I want a table with counts of age groups, with percents and cumulative counts/percents

desired_output <- 
  monkey_pox_agegroups %>% 
  #I want my calculations done per group, in this case agegroup
  group_by(agegroup) %>% 
  #this is a counting function, works greeeat with groupby
  tally(name = "Frequency") %>% 
  #let's create a new variable
  #ungrouping because I don't want this donee pere group now
  mutate(
    Percent = Frequency/sum(Frequency) * 100, #the sum function works like the SAS one,
    #and just, like the SAS a NULL value will NULL out the entire output, unless 
    #otherwise specificed to ignore NULLs or NAs
    
    
    #lets round this, this could be neested up there, but I like readability
    Percent = round(
      Percent,
      digits = 2
    ),
    
    cumulative_frequency = cumsum(Frequency),  #cumulative sum function
    
    #nesting here as an example
    cumulative_percent = round(
      cumulative_frequency / 
        sum(cumulative_frequency) * 100,
      digits = 2
    )
  )

desired_output
```

There is more or less your proc freq output.  Something like this could be tidied up
and stored into a function you can use over and over again, or another package could
be used with similar functions etc.... that will be out of the scope of this
project however.

For now let's pretty that table up to showcase some of the other dplyr verbs like
select, filter, rename and arrange.


```{r, prettyitup}
pretty_output <- 
  desired_output %>% 
  # performance wise, it's often best to filter at the top, lets drop rows
  #frequency is less than 20
  filter(Frequency >= 20) %>%  #any row where that statement is TRUE, will be kept
  #lets just select the first few columns
  select(
    agegroup,
    Frequency,
    Percent
  ) %>% 
  #lets rename agegroup 
  rename(
    #format is desired name first, then original var name
    # use backticks ` for non standard names
    `Age Group` = agegroup
  ) %>% 
  #lets arrange by highest frequeny in the descending ordere
  arrange(
    desc(Frequency)
  )
  
pretty_output
```

## Question

Based on the reading above, can you use the _summarise_ function, and the dataframe: 
monkey_pox_state_data to tell me which is the least common 'reportperiodend' date?

# Influenza 

## Overview

Influenza data will be going live to the very tables we have been working with in 
just a few days.  Respiratory Syncytial Virus(RSV) will likely follow suit, but that
is more up in the air.  Covid has shifted the seasonal patterns of other respiratory 
diseases and now many in the public health community are worried about the Covid, Flu
and RSV coming together to form a "tri-demic". 

Using what we've learned about R, visualizations, the ODP and github we're going to 
begin to put together a Rmarkdown document (like this one here) that will contain 
exploratory analyses surrounding ODP data for covid and flu that could be potentially
useful to public health decision makers. We will also attempt to incorporate signal
detection methods to see if there is any activity that might be out of the ordinary
and eventually we will take our findings and distill it down into key takeaways that
you might try to share with decision makers and stakeholders.


### Steps
* Review of ODP datasets for Covid-19 data (hint, there is a set of new tables and
a set of old tables)
* Review the ODP datasets for influenza data and any other publicly available data
  * If needed, engage with ODP data stewards from the Office of Policy and Management
  they can also help us comb through all the data to find useful datasets
* Construct a .rmd with possibly meaningful analyses
  * we will code what we can, but if there is anything you've learned biostats that 
  you think might be appropriate, I can work with you to employ those methods and fill
  in the 'r knowledge gaps'
* TBD
